The Effects of weather on maize yield: New evidence from Kenya

### 2018-12-26 writing/draft3/draftMonika.tex
writing

\address{Department of Geography, University of Sussex, Falmer, UK}
\ead{monika.novac@gmail.com}
\vspace{10pt}
\begin{indented}
\item[]November 2018
\end{indented}

@@

Following the procedure described above, we calculated a number of previously used characteristics of precipitation and temperature including indicators of floods. These measures are listed in table~\ref{vars} in the appendix. The measures which we found significant were: Total seasonal precipitation and its squared values, average seasonal temperature, coefficient of variation (CV) of seasonal precipitation, CV of average seasonal temperature (converted into Kelvin because the Celsius scale is interval and CV does not have any meaning for data measured on an interval scale),  maximum length of dry spell measured as number of consecutive dry days and the number of dry spells lasting for four days or more (for the purpose of this study we defined a dry day as a day when precipitation didn't exceed $1$mm). We did not include squared values of the average seasonal temperature in our preferred models as it was not significant and most of the past literature has not supported the evidence of the non-linear relationship between yield and temperature.
\subsection{Statistical approach}\label{stats}

@@

where $\mathbf{X}$ is an $n \times p$ model matrix of fixed effects, $\mathbf{\beta}$ is a $p$-dimensional fixed-effects parameter, $\mathbf{Z}$ is an $n \times q$ model matrix for the $q$-dimensional vector of random-effects variable $\mathscr{B}$ evaluated at $\mathbf{b}$ and $\sigma$ is a scale factor. The distribution of $\mathscr{B}$ can be written as:

@@

It was apparent from the histogram of the maize yield data that the distribution of maize yield was closer to a log-normal than to a normal distribution. Therefore, we opted for a log-linear functional form of our models as is a common practice in the agricultural economics. Furthermore, we scaled all predictors by subtracting mean and dividing them by standard deviation to avoid convergence problems.

To find the most suitable set of fixed effects, we adapted a stepwise selection procedure which minimized the Akaike's Information Criterion (AIC). The first step of the procedure was the estimation of the model with the complete set of our weather measures in the fixed effects and with the random intercepts. After this, we performed a search through the subsets of the fixed effects minimizing the AIC and allowing for steps in both `backward' and `forward' directions (that is removing and adding the predictors). We decided not to include number of heatwave days and average maximum daily temperature although including them as fixed effects would slightly improve the AIC. The reasons for not including them were their strong correlation with average seasonal temperature (the correlation coefficient was $0.981$ in the case of maximum daily temperature and $0.517$ for the number of heatwave days with the p-values smaller than $1\times10^{-6}$ in both cases), high variance inflation factor (VIF) of the maximum temperature ($5.397$) and the lack of significance of their individual t-tests (the p-values were $0.121$ for the maximum daily temperature and $0.075$ for the number of heatwave days). Furthermore, including heatwave days and maximum daily temperature would only lead to a small change in the AIC from $2080.877$ to $2076.439$ (these values differ from the AIC presented in table~\ref{ARMA} in the appendix because the final estimates reported in table~\ref{ARMA} were estimated using the restricted maximum likelihood (REML) whereas the models were re-estimated using the ML method for comparison\footnote{REML is generally the preferred estimation method for the mixed effects models as the ML estimates of the variance component are biased, however, the REML criterion is not meaningful for comparison of the models with different fixed effects. Therefore, we re-estimated the models by ML for the purpose of comparison \cite{Zuur2009}.}). In addition, the coefficients of both heatwave days and maximum daily temperature were positive, which is opposite to what we expected. Hence, we concluded that the decrease in the AIC and the positive signs are likely to be results of the high correlation with average temperature and we did not include them in our preferred specification.

We further applied a series of likelihood ratio tests to test for the significance of random slopes. No subset of the random slopes improved the fit; therefore, besides the fixed effects, our final model only included the random intercepts.\footnote{Although adding the random slope of CV improved the AIC from $2122.226$ to $2122.110$, we decided to not to include CV of temperature in the random effects. The reason for not including the random slope is that the difference in AIC was negligible and the F-test of the comparison of the two models was insignificant favouring the simpler model (p-value$=0.128$).}

To verify our results, we adapted an alternative step-down model building
approach using the Satterthwaite’s method to determine the p-values of the individual
t-tests \cite{lmerTest}. We started with a model which included the complete set of our weather measures in both fixed effects and random effects. In the first part of the procedure, the insignificant random effects were removed one by one. In each iteration, the insignificant variable with the highest p-value was removed. This was repeated until only the significant predictors remained. After this, we applied an analogous procedure to the fixed effects until we were left only with the significant variables. As recommended in literature, we considered the level of significance $\alpha=0.1$ for random effects and $\alpha=0.05$~ for fixed effects \cite{lmerTest}. This method led to the same model as the one obtained with our primary AIC-based method described above.

@@

Besides our main model, we estimated the same specification for two additional subsamples of the Kenyan counties. In particular, we estimated the model for the subsample of the ASAL counties and for the subsample of the non-ASAL counties separately to compare the effects of various weather characteristics across the two areas with the significantly different climate. The estimates of all three models are summarised in table~\ref{MainEst}. The first two columns represent the model for all counties, the third and fourth columns include estimates for the subsample of the ASAL counties and the last two columns describe the estimates based on the subsample of the non-ASAL counties. All predictors were standardised, therefore the estimates and the corresponding p-values can be interpreted as measures of relative importance. As an additional measure of relative importance we included F-values of the type \textit{III} analysis of variance (ANOVA) in table~\ref{MainEst}. The F-tests were conducted using the marginal rather than the sequential sum of squares. That is, for each variable, the F-test is a test of significance of that part of explained sum of squares which can be contributed to the particular explanatory variable. In other worlds, the F-value is a test statistic of comparison of the preferred model (with all the variables in table~\ref{MainEst}) and the model without the explanatory variable in question but including all other predictors. The main advantage of the type~\textit{III} ANOVA is that the test statistic does not depend on the order in which the effects are entered in the model.

The preferred specification has a log-linear functional form, therefore the estimates in table~\ref{MainEst} are not directly interpretable as the marginal effects and the coefficients are multiplicative. The predicted marginal effects of all coefficients are depicted in figures~\ref{MarEff1} in this section and in figures~\ref{MarEff2} and~\ref{MarEff3} in the appendix. The units of all predictors in all three figures are multiples of their standard deviations since the predictors were standardised. For the values of the marginal effects smaller than one, the effect on yield is negative while marginal effects higher than one can be interpreted as a positive effect. If, for example, a marginal effect of an explanatory variable $x$ was $1.1$ when $x$ is equal to its one standard deviation, one should interpret this as follows: if $x$ is equal to its standard deviation, the dependent variable (in our case maize yield) is $1.1$ times higher than it would be if the explanatory variable $x$ was equal to its mean (which is zero in figures~\ref{MarEff1} in this section and in figures~\ref{MarEff2} and~\ref{MarEff3} in the appendix as the predictors were standardised).

@@

@@

Other fixed effects which we found significant include CV of temperature and precipitation and length and number of dry spells (see the first column of table~\ref{MainEst}). As it is apparent from figures~\ref{MarEff2}(a) and \ref{MarEff2}(b) in the appendix, the relationship of maize yield and each of these variables is linear. In accordance with our expectation, the yield decreases with increasing CV of precipitation. \hl{However, the result which we found for CV of temperature is a bit puzzling as the yield seem to increase with increasing variability of temperature. This is probably because...}

@@

According to the magnitude of the standardised coefficients and F-values in the first two columns of table~\ref{MainEst}, the most important variable is average seasonal temperature which appears to be much more important than the remaining predictors. The second most important variable is total seasonal precipitation. The value of the F-statistic corresponding to temperature is $12.010$ while the F-statistic of the linear term of cumulative precipitation is equal to $5.358$ and the F-statistic of its quadratic term is $4.263$. Similarly, according to the Satterthwaite's p-values, average temperature is the most important predictor as the corresponding Satterthwaite's p-value is equal to $6\times10^{-4}$ which is much lower than the p-values of the other explanatory variables.

We will now discuss and compare the results of the models based on the subsample of the ASAL counties (the third and fourth columns of table~\ref{MainEst}) and the model based on the subsample of the non-ASAL counties (the fifth and sixth columns of table~\ref{MainEst}). Interestingly, different predictors were found significant for each of the two subsamples. Regarding the subsample of the ASAL counties, the only significant fixed effects are average seasonal temperature and length and number of dry spells. According to our expectation, all three variables have a negative sign. On the other hand, the variables which were found significant for the subsample of the non-ASAL counties include seasonal precipitation (positive) and its square (negative) and CV of temperature (positive). According to the F-statistics, the most important explanatory variables for the ASAL counties are number of dry spells and length of the longest dry spell with values of the F-statistics equal to $8.032$ and $6.841$, respectively. The F-statistic of the average temperature is $5.409$ and the F-statistics of the remaining predictors are lower than zero. However, according to the magnitude of the standardised coefficients, the average temperature would be the most important as the coefficient is equal to $-0.214$ while it is equal to $-0.156$ and $-0.181$ for the number of dry spells and the length of maximum dry spell, respectively. With regard to the subsample of the non-ASAL counties, the most important explanatory variables based on the F-statistic are precipitation and its square. The F-statistic of the squared values of precipitation is $23.384$ and its value for the linear term of precipitation is $19.140$, while the variable with the third highest F-statistic is CV of temperature with the F-statistic equal to $5.372$, hence, the difference in importance of the precipitation and CV of temperature is relatively large. The standardised coefficient of squared precipitation is equal to $-0.127$, the coefficient of the linear term of precipitation is $0.244$ and the standardised coefficient of CV of temperature is $0.058$. In terms of the magnitude of the coefficients, the average temperature is also relatively important as the value of the coefficient is $-0.128$. However, average temperature is insignificant according to the Satterthwaite's p-value and its F-statistic is also relatively small, in particular $1.528$.

To sum up, for the ASAL counties temperature is much more important than for the non-ASAL counties where temperature is not significant. On the other hand, precipitation is not significant for the ASAL counties while it is strongly significant for the non-ASAL counties. This is illustrated in figure~\ref{MarEff1}(c)-(f) in this section. The plot of the marginal effect of precipitation is almost flat for the ASAL counties (figure \ref{MarEff1}(c) in this section) while it is hill-shaped and steep for the non-ASAL counties (figure \ref{MarEff1}(e) in this section) similarly as in the case of the whole sample. As one can see in figures~\ref{MarEff1}(d) and~\ref{MarEff1}(f) in this section, the marginal effect of temperature is linear and decreasing for the case of ASAL and non-ASAL counties. Marginal effects of the remaining predictors are depicted in figures~\ref{MarEff2}(c)-(f) and~\ref{MarEff3}(c)-(f) in the appendix.

@@

The AIC is an important criterion for comparison of competing models, however it does not provide any information about the absolute goodness of fit. In order to assess the absolute goodness of fit of a model, a different measure is needed. For a simple linear regression, the coeffiecient of determination $R^2$ or its adjusted variant is typically used for this purpose. Nevertheless, obtaining the $R^2$ or the adjusted $R^2$ is not straightforward for linear mixed models or for models estimated by the maximum likelihood in general. In this study, we used the coefficient of determination of generalized linear mixed-effects models introduced by \cite{Nakagawa2013}. For each of our samples (all counties, ASAL and non-ASAL), we present two versions of the statistic in table~\ref{R2}. These include marginal $R^2$ which can be interpreted as the variance explained by the fixed effects components and the conditional $R^2$ which can be interpreted as the variance explained by both fixed effects and random effects. In this case, the only random effects are the intercepts, hence the difference between the two variants is the variance explained by the random intercepts. As is apparent from table~\ref{R2}, the fixed effects explain about $30\%$ of the variance in case of the model for all counties.

\begin{center}
\begin{threeparttable}
\caption{\label{R2}Coefficient of determination $R^2$ for mixed models}
\begin{indented}
\item[]
\begin{tabular}{@{}lcc}
\\[-1em]
\br
\\[-1em]
\textbf{Sample:}&Marginal\tnote{a}&Conditional\tnote{b}\\
\hline
{All counties}& $0.300$&$0.526$\\
{ASAL}& $0.134$&$0.353$\\
{non-ASAL}&  $0.153$&$0.436$\\
\br
\end{tabular}
\begin{tablenotes}
\begin{footnotesize}
    \item \textit{Notes:} The coefficient of determination from generalised linear mixed-effects models. The statistic is based on \cite{Nakagawa2013}.  \\
        \begin{adjustwidth}{1cm}{}
    \item[a] The marginal coefficient is interpreted as the variance explained by the fixed effects
  \item[b] while the conditional one is explained by both fixed affects and random effects (in our case by fixed effects and random intercepts
     \end{adjustwidth}
     \end{footnotesize}
\end{tablenotes}

@@

To check for presence of multicollinearity we calculated the variance inflation factors (VIF) of the explanatory variables. With only two exceptions, the VIF were below $2$ for all predictors in our final models. For CV of precipitation, the value of VIF was $2.172$ in the model of all counties, $2.147$ in the model for the subsample of the ASAL counties and $2.182$ in the model for the subsample of the non-ASAL counties. These values are still well below the threshold of problematic multicollinearity. The other exception was the seasonal precipitation as its VIF is $2.407$ in the model for the non-ASAL counties, probably as a result of the presence of the squared of precipitation. The VIF is below $2$ in the other two models and even the value $2.407$ is below the threshold of the problematic multicollinearity. Therefore, we can conclude that the models do not suffer from presence of multicollinearity.

\begin{itemize}
\color{blue}
\item Verbal description and interpretation of the results. Discussing goodness of fit using various criteria such as AIC or alternatives to $R^2$.

\item relative importance of the individual variables and its measures
%\item Possibly estimate simple regressions with the same explanatory variables as the preferred mixed effects specification. This would give us an approximation of $R^2$ which could be used to compare our models.

@@

\item show VIF
\item If we get the yield data for the period from $2015$ onwards: Out of sample predictions and comparison with the real data
\end{itemize}

@@

\vspace{1cm}
\textcolor{blue}{Possibly include a table of all values which I get from the lme or lme4 models summary in R, that is. correlation of the fixed effects etcetera}

\textcolor{blue}{Maybe also a table with standard errors here}

>>>>

\address{Department of Geography, University of Sussex, Falmer, UK}
\ead{monika.novac@gmail.com}
\vspace{10pt}
\begin{indented}
\item[]December 2018
\end{indented}

@@

Following the procedure described above, we calculated a number of previously used characteristics of precipitation and temperature including indicators of floods. These measures are listed in table~\ref{vars} in the appendix. The measures which we found significant were: Total seasonal precipitation and its squared values, average seasonal temperature, coefficient of variation (CV) of seasonal precipitation, CV of average seasonal temperature (converted into Kelvin because the Celsius scale is interval and CV does not have any meaning for data measured on an interval scale),  maximum length of dry spell measured as number of consecutive dry days and the number of dry spells lasting for four days or more (for the purpose of this study we defined a dry day as a day when precipitation didn't exceed $1$mm). We did not include squared values of the average seasonal temperature in our preferred models as it was not significant and most of the past literature has not supported the evidence of the non-linear relationship between yield and average temperature.
\subsection{Statistical approach}\label{stats}

@@

where $\mathbf{X}$ is an $n \times p$ model matrix of fixed effects, $\mathbf{\beta}$ is a $p$-dimensional fixed-effects parameter, $\mathbf{Z}$ is an $n \times q$ model matrix for the $q$-dimensional vector of random-effects variables $\mathscr{B}$ evaluated at $\mathbf{b}$ and $\sigma$ is a scale factor. The distribution of $\mathscr{B}$ can be written as:

@@

It was apparent from the histogram of the maize yield that the distribution of the yield data was closer to a log-normal than to a normal distribution. Therefore, we opted for a log-linear functional form of our models as is a common practice in the agricultural economics. Furthermore, we scaled all predictors by subtracting their mean and dividing them by their standard deviation to avoid convergence problems during estimation procedures.

To find the most suitable set of fixed effects, we adapted a stepwise selection procedure which minimized the Akaike's Information Criterion (AIC). The first step of the procedure was the estimation of the model with the complete set of our weather measures in the fixed effects and with the random intercepts. After this, we performed a search through the subsets of the fixed effects minimizing the AIC and allowing for steps in both `backward' and `forward' directions (that is removing and adding the predictors). We decided not to include number of heatwave days and average maximum daily temperature although including them as fixed effects would slightly improve the AIC. The reasons for not including them were their strong correlation with average seasonal temperature (the correlation coefficient was $0.981$ in the case of maximum daily temperature and $0.517$ for the number of heatwave days with the p-values smaller than $1\times10^{-6}$ in both cases), high variance inflation factor (VIF) of the maximum temperature ($5.397$) and the lack of significance of their individual t-tests (the p-values were $0.121$ for the maximum daily temperature and $0.075$ for the number of heatwave days). Furthermore, including heatwave days and maximum daily temperature would only lead to a small change in the AIC from $2080.877$ to $2076.439$ (these values differ from the AIC presented in table~\ref{ARMA} in the appendix because the final estimates reported in table~\ref{ARMA} were obtained using the restricted maximum likelihood (REML) whereas the models were re-estimated using the ML method for the comparison\footnote{REML is generally the preferred estimation method for the mixed effects models as the ML estimates of the variance component are biased, however, the REML criterion is not meaningful for comparison of the models with different fixed effects. Therefore, we re-estimated the models by ML for the purpose of comparison \cite{Zuur2009}.}). In addition, the coefficients of both heatwave days and maximum daily temperature were positive, which is opposite to what we expected. Hence, we concluded that the decrease in the AIC and the positive signs are likely to be results of the high correlation with average temperature and we did not include them in our preferred specification.

After we found a set of significant fixed effects, we applied a series of likelihood ratio tests to test for the significance of random slopes. No subset of the random slopes improved the fit; therefore, besides the fixed effects, our final model only includes the random intercepts.\footnote{Although adding the random slope of CV of temperature improved the AIC from $2122.226$ to $2122.110$, we decided to not to include this variable in the random effects. The reason for not including the random slope is that the difference in AIC was negligible and the F-test of the comparison of the two models was insignificant favouring the simpler model (p-value$=0.128$).}

To verify our results, we adapted an alternative step-down model building
approach using the Satterthwaite’s method to determine the p-values of the individual
t-tests \cite{lmerTest}. We started with a model which included the complete set of our weather measures in both fixed effects and random effects. In the first part of the procedure, the insignificant random effects were removed one by one. In each iteration, the insignificant variable with the highest p-value was removed. This was repeated until only the significant random effects remained. After this, we applied an analogous procedure to the fixed effects until we were left only with the significant variables. As recommended in literature, we considered the level of significance $\alpha=0.1$ for the random effects and $\alpha=0.05$~ for the fixed effects \cite{lmerTest}. This method led to the same model as the one obtained with our primary AIC-based method described above.

@@

Besides our main model estimated for all counties of Kenya, we estimated the same specification for two subsamples of the Kenyan counties. In particular, we estimated the model for the subsample of the ASAL counties and for the subsample of the non-ASAL counties separately to compare the effects of various weather characteristics across the two areas with significantly different climate. The estimates of all three models are summarised in table~\ref{MainEst}. The first two columns represent the model for all counties, the third and fourth columns include estimates for the subsample of the ASAL counties and the last two columns describe the estimates based on the subsample of the non-ASAL counties. All predictors were standardised, therefore the estimates and the corresponding p-values can be interpreted as measures of relative importance. As an additional measure of relative importance we included F-values of the type \textit{III} analysis of variance (ANOVA) in table~\ref{MainEst}. The type \textit{III} ANOVA is conducted using the marginal rather than the sequential sum of squares. That is, for each variable, the F-test is a test of significance of that part of explained sum of squares which can be contributed to the particular explanatory variable. In other worlds, the F-value is a test statistic of comparison of the preferred model (with all the variables in table~\ref{MainEst} in our case) and the model without the explanatory variable in question but including all other predictors. The main advantage of the type~\textit{III} ANOVA is that the test statistic does not depend on the order in which the effects are entered in the model.

The preferred specification has a log-linear functional form, therefore the estimates in table~\ref{MainEst} are not directly interpretable as marginal effects and the coefficients are multiplicative. The predicted marginal effects of all coefficients are depicted in figures~\ref{MarEff1} in this section and in figures~\ref{MarEff2} and~\ref{MarEff3} in the appendix. The units of all predictors in all three figures are multiples of their standard deviations since the predictors were standardised. For the values of the marginal effects smaller than one, the effect on yield is negative while marginal effects higher than one can be interpreted as a positive effect. If, for example, a marginal effect of an explanatory variable $x$ being equal to its standard deviation is $1.1$, one should interpret this as follows: if $x$ is equal to its standard deviation, the dependent variable (in our case maize yield) is $1.1$ times higher than it would be if the explanatory variable $x$ was equal to its mean (which is zero in figure~\ref{MarEff1} in this section and in figures~\ref{MarEff2} and~\ref{MarEff3} in the appendix as the predictors were standardised).

@@

\vspace{1cm}

@@

Other fixed effects which we found significant include CV of temperature and precipitation and length and number of dry spells (see the first column of table~\ref{MainEst}). As it is apparent from figures~\ref{MarEff2}(a)-(b) and \ref{MarEff3}(a)-(b) in the appendix, the relationship of maize yield and each of these variables is linear. In accordance with our expectation, the yield decreases with increasing CV of precipitation. \hl{However, the yield seems to increase with increasing variability of temperature which is counter-intuitive...}\textcolor{blue}{...\textit{reasoning}...}

@@

According to the magnitude of the standardised regression coefficients and F-values in the first two columns of table~\ref{MainEst}, the most important variable is average seasonal temperature which appears to be much more important than the remaining predictors. The second most important variable is total seasonal precipitation. The value of the F-statistic corresponding to temperature is $12.010$ while the F-statistic of the linear term of cumulative precipitation is equal to $5.358$ and the F-statistic of its quadratic term is $4.263$. Similarly, according to the Satterthwaite's p-values, average temperature is the most important predictor as the corresponding Satterthwaite's p-value is equal to $6\times10^{-4}$ which is much lower than the p-values of the other explanatory variables.

We will now discuss and compare the results of the models estimated for the subsample of the ASAL counties (the third and fourth columns of table~\ref{MainEst}) and the model estimated for the subsample of the non-ASAL counties (the fifth and sixth columns of table~\ref{MainEst}). Interestingly, different predictors were found significant for each of the two subsamples. Regarding the subsample of the ASAL counties, the only significant fixed effects are average seasonal temperature and length and number of dry spells. According to our expectation, all three variables have a negative sign. On the other hand, the variables which were found significant for the subsample of the non-ASAL counties are seasonal precipitation (positive) and its square (negative) and CV of temperature (positive). According to the F-statistics, the most important explanatory variables for the ASAL counties are number of dry spells and length of the longest dry spell with values of the F-statistics equal to $8.032$ and $6.841$, respectively. The F-statistic of the average temperature is $5.409$ and the F-statistics of the remaining predictors are lower than one. However, according to the magnitude of the standardised regression coefficients, the average temperature is the most important as its coefficient is equal to $-0.214$ while it is equal to $-0.156$ and $-0.181$ for the number of dry spells and the length of maximum dry spell, respectively.

With regard to the subsample of the non-ASAL counties, the most important explanatory variables based on the F-statistic are precipitation and its squared term. The F-statistic of the squared values of precipitation is $23.384$ and the F-value of the linear term of precipitation is $19.140$. The variable with the third highest value of the F-statistic is CV of temperature and the corresponding F-statistic is equal to $5.372$. Hence, the difference in importance of the precipitation and CV of temperature is relatively large. The standardised regression coefficient of squared precipitation is equal to $-0.127$, the coefficient of the linear term of precipitation is $0.244$ and the standardised regression coefficient of CV of temperature is $0.058$. In terms of the magnitude of the coefficients, the average temperature is also relatively important as the value of its coefficient is $-0.128$. However, average temperature is insignificant according to the Satterthwaite's p-value and also the value of its F-statistic is relatively low ($1.528$).

To sum up, temperature is much more important for the ASAL counties than for the non-ASAL counties where temperature is isignificant. On the other hand, precipitation is not significant for the ASAL counties while it is strongly significant for the non-ASAL counties. This is illustrated in figures~\ref{MarEff1}(c)-(f) in this section. The plot of the marginal effect of precipitation is almost flat for the ASAL counties (figure \ref{MarEff1}(c) in this section) while it is hill-shaped and steep for the non-ASAL counties (figure \ref{MarEff1}(e) in this section) similarly as in the case of the whole sample. As one can see in figures~\ref{MarEff1}(d) and~\ref{MarEff1}(f) in this section, the marginal effect of temperature is linear and decreasing for the both subsamples . Marginal effects of the remaining predictors are depicted in figures~\ref{MarEff2}(c)-(f) and~\ref{MarEff3}(c)-(f) in the appendix.

@@

The AIC is an important criterion for comparison of competing models, however it does not provide any information about the absolute goodness of fit. In order to assess the absolute goodness of fit of a model, a different measure is needed. For a simple linear regression, the coeffiecient of determination $R^2$ or its adjusted version is typically used for this purpose. Nevertheless, obtaining the $R^2$ or the adjusted $R^2$ is not straightforward for linear mixed models or for models estimated by the maximum likelihood in general. In this study, we used the coefficient of determination of generalized linear mixed-effects models introduced by \cite{Nakagawa2013}. For each of our samples (all counties, ASAL and non-ASAL), we present two versions of the statistic in table~\ref{R2}. These include marginal $R^2$ which can be interpreted as the variance explained by the fixed effects and the conditional $R^2$ which can be interpreted as the variance explained by both fixed effects and random effects. In our case, the only random effects are the intercepts, hence the difference between the two variants is the variance explained by the random intercepts. As is apparent from table~\ref{R2}, the fixed effects explain about $30\%$ of the variance in case of the model for all counties,  about $13\%$ of the variance in the model based on the ASAL counties and about $15\%$ of the variance in the model based on the non-ASAL counties.

\begin{center}
\begin{threeparttable}
\caption{\label{R2}Coefficient of determination $R^2$ for mixed models}
\begin{indented}
\item[]
\begin{tabular}{@{}lcc}
\\[-1em]
\br
\\[-1em]
\textbf{Sample:}&Marginal\tnote{a}&Conditional\tnote{b}\\
\hline
{All counties}& $0.300$&$0.526$\\
{ASAL}& $0.134$&$0.353$\\
{non-ASAL}&  $0.153$&$0.436$\\
\br
\end{tabular}
\begin{tablenotes}
\begin{footnotesize}
    \item \textit{Notes:} The coefficient of determination from generalised linear mixed-effects models. The statistic is based on \cite{Nakagawa2013}.  \\
        \begin{adjustwidth}{1cm}{}
    \item[a] The marginal $R^2$ is interpreted as the variance explained by the fixed effects
  \item[b] The conditional $R^2$ is explained by both fixed affects and random effects (in our case by fixed effects and random intercepts)
     \end{adjustwidth}
     \end{footnotesize}
\end{tablenotes}

@@

Following a standard practise, we calculated the variance inflation factors (VIF) of the explanatory variables to test for presence of multicollinearity. With only two exceptions, the VIF is below $2$ for all predictors in our final models. For CV of precipitation, the value of VIF is $2.172$ in the model of all counties, $2.147$ in the model for the subsample of the ASAL counties and it is equal to $2.182$ in the model based on the subsample of the non-ASAL counties. These values are well below the threshold of problematic multicollinearity. The other exception is the linear term of seasonal precipitation as its VIF is $2.407$ in the model for the non-ASAL counties, probably due to presence of the squared term of precipitation. However, $2.407$ is below the threshold of the problematic multicollinearity and the VIF of precipitation is below $2$ in the other two models. Therefore, we can conclude that the models do not suffer from presence of multicollinearity.

%\item Possibly estimate simple regressions with the same explanatory variables as the preferred mixed effects specification. This would give us an approximation of $R^2$ which could be used to compare our models.

@@

\begin{itemize}
\color{blue}
\item If we get the yield data for the period from $2015$ onwards: Out of sample predictions and comparison with the real data
\end{itemize}

@@

\vspace{1cm}
%\textcolor{blue}{Possibly include a table of all values which I get from the lme or lme4 models summary in R, that is. correlation of the fixed effects etcetera}

%\textcolor{blue}{Maybe also a table with standard errors here}
